---
title: "final_template"
author: "Hibah Ganie"
date: "2025-06-10"
output: html_document
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---
## [Hibah's Github Repository](https://github.com/hibdawg/ENVS-193DS_spring-2025_final)

```{r Reading In Packages}
library(tidyverse) # General use
library(janitor) # Cleaning data Frames
library(dplyr) # Mutate columns
library(lubridate) # make dealing with dates easier
library(here) # File Organization
library(readxl) # reading .xlsx files
library(ggeffects) # generating model predictions
library(gtsummary) # generating summary tables for models 
library(ggplot2) # visualizations
library(hms) # working with times
```

```{r Reading In Data}
sst <- read_csv(here("data", "SST_update2023.csv")) # read in my data
```

# Problem 1. Research Writing

## a. 
[In part one they used a correlation test to determine how two variables, distance from headwater (km) and annual total nitrogen load (kg year^-1) relate or are associated with eachother. My coworker also use an ANOVA in part 2 to compare the means of more than two groups.]

## b.
[Accepting uncertainty is key in interpreting analysis of variance, and can be conducted through communicating the confidence intervals, such as the CI of the difference between urban land, atmospheric desposition, fertilizer, wastewater treatment, and grasslands.Tukey's Honestly Signifiant difference can be used to further investigate which groups are diffent when compares, giving information on pairwise comparison, difference in means between groups, and adjusted p-value to increase transparency.Effect size can also be used to determine how strongly distance from headwater (KM) effects the annual total nitrogen load (kg year^-1) found between groups.]

 
## c. 
[We found a neglible difference (n^2= effect size) between groups in mean nitrogen load (One-Way Anova, F-distribution(among groups Degrees of Freedom,within group Degrees of Freedom) = f-value, p < 0.02, $\alpha$ = signifcance level)

On average, there was little difference of nitrogen load between urban land, atmospheric deposition, fertilizer, wastewater treatment and grasslands (Tukey HSD difference, % Confidence Interval:[CI, CI](kg year^-1) based on distance from headwater (KM)) ]

# Problem 2. Data Visualization

## a. Cleaning and Summarizing 

```{r Cleaning SST Data}
sst_clean <- sst |> 
  # select sst data frame
  clean_names() |> 
  # remove spaces and uppercase letters (cleans)
  mutate( # mutate function to change formatting of date, year, and month using lubridate package
    date = ymd(date), #year, month, date formate for date column                    
    year = year(date), # change from chr to num for year  
    month = month(date, label = TRUE, abbr = TRUE) # change month to an ordered factor and abbreviate
  ) |> 
  filter(year >= 2018) |> 
  # filters years to after 2018
  mutate(
    year = factor(year), # convert year to factor
    month = factor(month, levels = month.abb, ordered = TRUE) # month as a factor
  ) |> 
  group_by(year, month) |> 
  # group together year and month for mean monthly SST calculation
  summarise(mean_monthly_sst = 
              mean(temp, na.rm = TRUE),
            .groups = "drop") |> 
  # Calculate mean monthly sst and ignore missing values, then drop the grouping
  mutate(
    mean_monthly_sst = (format(round(mean_monthly_sst, 1), nsmall = 1) # round values to one decimal point
  )) |> 
  # One last mutate to convert from chr to number, or dbl format
  mutate( 
    mean_monthly_sst = 
      as.numeric(mean_monthly_sst)
  )

# Randomly display 5 rows from sst_clean data frame
sst_clean |> 
  slice_sample(n = 5)

#display structure of dataframe
str(sst_clean)
```

```{r Creating Time-Series Visualization}

ggplot(data = sst_clean, 
       aes(x=month, 
           y=mean_monthly_sst,
           group = year,
           color= factor(year))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(
    values= c(
      "2018" = "#aec7e8",
      "2019" = "#7fc6e8",
      "2020" = "#487db1",
      "2021" = "#2c4d79",
      "2022" = "#1e2f40", 
      "2023" = "#000000"
      )) +
  labs(x = substitute(paste(bold('Month'))),
       y = substitute(paste(bold("Mean Monthly Surface Temperature (°C)"))),
       color = substitute(paste(bold("Year"))),
       ) +
  theme(
    panel.background = element_rect(
      fill = "white",
      colour = NA),       # white panel fill, no border here
    panel.border = element_rect(
      fill = NA, 
      colour = "#808080", 
      size = 1),  # thick black border
    plot.background  = element_rect(fill = "white", colour = NA),       # white outside panel
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks        = element_blank()) +
  theme(
    legend.position = c(.16,.95), # adjust legend position to top left inside the graph
    legend.justification = c(.8,1) # make corner of legend to corner of graph
  )
```














